Geometric variational problems

### 2020-11-02 NSF2020/Description/specificaim2_ver2.tex
working on my section of proposal

\subsection{Specific Aim 2: Efficient high-order numerical algorithm for large-scale simulation of the HAP model}\label{subsec:specific_aim_2}
% -----------------------------------------------------------------------------
% {{{
\subsubsection{Overview of the potential theory and boundary integral equation methods}
The HAP model requires solving the exterior Dirichlet problem of the screened Laplace
equation and the mobility problem of the Stokes flow at each time step.
Standard numerical methods such as finite difference and finite element
methods have to truncate the infinite domain to a finite
computational domain by imposing certain artificial boundary conditions
on the truncated domain and need the discretization of the whole volume
of the truncated domain. The artificial boundary conditions are very often
inexact, introducing another layer of approximation that
lowers the accuracy of the solution. The discretization of the whole volume is
very expensive
for three dimensional problems, leading to excessively large number of unknowns in
the linear system. Furthermore, it is rather difficult to obtain high-order
discretization when the boundary surfaces are irregular.

@@

The BIE method removes the need of imposing artificial boundary conditions
for exterior problems
and reduces the dimension of the problem by one, leading to optimal number of unknowns
in the solve phase. Very often, one is able to construct the so-called
{\it second kind integral equation} (SKIE) formulation for these problems,
where the operator is a sum of the identity operator and a compact operator.
It is well-known that the condition number of the
resulting linear system for SKIEs is {\it independent} of the mesh size $h$ or equivalently,
the number of discretization points. This leads to
a constant number of iterations when iterative solvers are applied to solve
the resulting linear system. It further
improves the accuracy of the solution as the relative error of solution
will never increase as the number of total discretization points increases.
On the other hand, the condition
number of the linear system from finite difference and finite element methods generally
increases as a power function of the total number of dicretization points, with
the exact power depending on the underlying spatial dimension and the highest
order of the differential operator in the PDE. This leads to an increase in the number
of iterations (and hence the total computational cost) and an increase in the relative
error of the solution as the number of dicretization points increases.

The discretization of the integral operator often leads to a dense matrix,
while finite difference/finite element methods result in sparse matrices,
making matrix-vector product much cheaper to compute. However,since the invention
of the original fast multipole method (FMM)\cite{fmm5}, there
have been many fast algorithms~\cite{fmm1,fmm2,fmm3,fmm4,fmm6,fmm7,fmm8}
that reduce the cost of matrix-vector product
to linear ($O(N)$) or quasilinear complexity ($O(N \log^kN)$) for dense matrices resulting
from the BIE discretization.
Fast direct solvers have also been developed
recently~\cite{fds1,fds2,fds3,fds4,fds5,fds6,fds7,fds8,ho2016cpam2,ho2016cpam1,minden2016,minden2017siammms}.
The development of these fast algorithms have removed a huge
hurdle for the use of boundary integral equations, making the BIE method the method
of choice for constant-coefficient PDEs with linear boundary conditions. For the exterior
Dirichlet problem of the screened Laplace equation, the SKIE formulation in our previous
work~\cite{Fu2018_SIAM} can be readily extened to the 3D case. For the mobility problem
of the Stokes flow in three dimensions that
takes the hydrodynamic effect into account, an SKIE formulation
can be found in \cite{manasthesis}.

\subsubsection{High-order discretization of surface integrals in three dimensions}
The practical application of integral equation methods requires the
accurate evaluation of boundary integrals with singular, weakly
singular or nearly singular kernels. There are many numerical quadrature schemes
for dealing with line integrals in two dimensions~\cite{alpert,kapur,sidi,duffy,bruno1,bruno2,davis_1984,graglia_2008,hackbusch_sauter_1994, jarvenpaa_2003,khayat_2005,kress_boundary_1991,schwab_1992, ying_2006,beale1,beale2,goodman_1990, haroldson_1998, lowengrub_1993,schwab_1992,ggq1,ggq2,ggq3,helsing_2008a,helsing_integral_2009,helsing_tutorial_2012}
and some of them are extremely efficient and can achieve arbitrary high order.

The high-order quadrature rules for the evaluation of surface integrals
in three dimensions, however, are much less developed than the line integrals in two
dimensions. For example, there are no Gaussian quadratures for integraing
polynomials on a flat triangle, even though efficient quadratures
\cite{xiao2010cma,vioreanu2014} have been
developed recently for such purpose. For weakly singular or singular integrals,
\cite{bremer2012jcp,bremer2013jcp} constructed high-order quadratures
for surface integrals on a general triangle, while \cite{gimbutas2013sisc}
presented a fast algorithm for integrating $1/r$-type singular integrals
for surfaces that are homeomorphic to a sphere. We plan
to study the so-called quadrature by expansion (QBX) scheme~\cite{klockner2013jcp,qbx2}
for the evaluation
of both singular and near-singular surface integrals encountered in the
discretization of BIEs in three dimensions. Conceptually, the idea of the QBX
to evaluate singular, hypersingular and near singular integrals
on smooth surfaces is straightforward. That is, the surface is discretized
into smooth triangles and smooth high-order quadratures are applied to evaluate
the expansion coefficients
on all source triangles with the QBX expansion center placed at a point off the surface.
One may then form a suitable expansion (for example, a Taylor expansion) around that
center and evaluate this expansion back at the target point on the surface (or close
to the surface in the near singular case). Compared to the competing aforementioned
quadrature schemes, the QBX quadrature is attractive
because it offers a clear path for being extended to: \textbf{(1)}
handle any singularity, including
hypersingular operators, \textbf{(2)} be usable with any high-order surface
discretization, \textbf{(3)} generate well-conditioned discrete
operators to which iterative methods such as GMRES~\cite{gmres} can be
applied in a black-box fashion, \textbf{(4)} integrate well with fast algorithms
such as the FMM.

In practice, there are still many issues that need to be resolved.
For example, there are now many variants of QBX including global and local
QBX~\cite{klockner2013jcp,rachh2017jcp}, the target-specific QBX~\cite{siegel2018jcp},
kernel-independent QBX~\cite{abtin2018bit}, and
quadrature by two expansions~\cite{ding2019arxiv}. The coupling of the QBX
and the FMM may also lead to certain instability issues which may require
some changes in the fast multipole method~\cite{wala2018jcp}. Similar
to other quadrature methods, there have been extensive study on the QBX
methods in two dimensions, while its three dimension
version~\cite{wala2019jcp,af2018sisc,siegel2018jcp,wala2019arxiv} has not been
fully studied and the implementation is even more scarce. We plan to investigate
the accuracy and the convergence order of the various QBX schemes mentioned above,
its coupling with the FMM, parallel implementation issues for large-scale
problems, and the application to our target problems.
\subsubsection{A hybrid method for close-to-touching Janus particles}
The HAP model for membranes will inevitably lead to the case where some Janus particles
will be very close to each other. Even though QBX can handle such case
accurately, it may lead to excessively large number of unknowns in order to resolve
the unknown densities at those close-to-touching points. In \cite{gan2016sisc},
a hybrid scheme is developed to treat the close-to-touching spheres for the interface
problem of the Laplace equation. The key idea is to combine the analytic image method
and spectrally accurate method of moments, where the analytic image method will capture
the most singular part of the unknown density at the close-to-touching points.
With the aid of the FMM, the scheme achieves linear complexity with much less number
of unknowns on the material interfaces.
We plan to study the hybrid scheme
to treat the close-to-touching case for both the screened Laplace equation and the
mobility problem to further improve the efficiency of the numerical algorithm
for the HAP model.
\subsubsection{Other numerical issues}
Another outstanding numerical issue pertaining to simulating the self-assembly of amphiphilic particles (such as lipid macromolecles) in a
viscous solvent is the collision between amphiphilic particles.
The hydrophobic attraction potential drives the amphiphilic particles to move towards each other so to minimize exposure to the solvent. 
Under such
attraction force the fluid is being squeezed out as particles move toward each other, and the amphiphilic particles can come into physical
contact in finite time. 
Such particle collisions in a dense rigid body suspension is a great challenge and can be a bottleneck in large-scale simulations. 
In our previous work a Leonard-Jones potential is used to keep the
particles from overlapping \cite{Fu2018_SIAM}. 
Such steep steric interaction at
short ranges introduces great numerical stiffness that limit the time step size.

Yan {\it et al.} proposed
a collision-resolution algorithm within the framework of boundary integral methods to resolve the non-smooth many-body
dynamics due to collisions by formulating the particle collision process as a linear complementarity problem with geometric 
`non-overlapping' constraint \cite{Yan2019}. 
We propose to treat the particle collisions based on Yan {\it et al.}'s work after the first two goals (see above)
are achieved. 
A potential complication in combining the algorithms for particle collision into our 
QBX integral formulation is the computational cost for the QBX to refine meshes on particles in collision.

Once the collision-resolution algorithm is implemented, we will check the computing performance first before we embark on 
incorporating fluctuating hydrodynamics \cite{Bao17,Bao18}, which might be important even at the scales of the amphiphilic 
particle size in some cases. We propose to extend the high-order
time-integration scheme for a stochastic differential equation
\cite{fu2015pre} to
this system.

>>>>

\subsection{Specific Aim 2: Efficient high-order numerical algorithm for
large-scale simulation of the HAP model}
\label{subsec:specific_aim_2}
% ----------------------------------------------------------------------
\subsubsection{Overview of the potential theory and boundary integral
equation methods} 
The HAP model requires solving the exterior Dirichlet problem of the
screened Laplace equation and the mobility problem of the Stokes flow at
each time step. If discretizing these equations with stencil-based
numerical methods, the computational domain must be truncated, the
volume must be discretized, and artificial boundary conditions must be
imposed. The artificial boundary conditions introduces an additional
source of error, and discretizing the volume, particularly in three
dimensions, results in an excessively large linear system. Moreover, it
is difficult to obtain high-order discretizations when the boundary
surfaces are irregular. Since we are solving constant-coefficient
elliptic PDEs, we apply boundary integral equation (BIE) that remove the
aforementioned obstacles. BIEs represent the solution as a layer
potential with an unknown density function. Then, the correct density
function is found that matching the layer potential with the prescribed
boundary conditions. By construction, layer potentials satisfy the
underlying PDE and the far-field conditions.

Discretizations of BIEs have several advantages when compared to
stencil-based discretizations. First, the dimension of the problem is
reduced by one since the unknown density function is defined on the
boundary of the geometry. Second, by carefully selecting each layer
potential, the resulting linear system is well-conditioned for any
resolution and iterative methods such as GMRES require a
mesh-independent number of iterations to converge. Third, by using
appropriate quadrature methods, the method converges with spectral or
very high-order accuracy.

Challenges are dense matrices, preconditioners, and near-singular
integration

%The HAP model requires solving the exterior Dirichlet problem of the
%screened Laplace equation and the mobility problem of the Stokes flow
%at each time step. Standard numerical methods such as finite difference
%and finite element methods have to truncate the infinite domain to a
%finite computational domain by imposing certain artificial boundary
%conditions on the truncated domain and need the discretization of the
%whole volume of the truncated domain. The artificial boundary
%conditions are very often inexact, introducing another layer of
%approximation that lowers the accuracy of the solution. The
%discretization of the whole volume is very expensive for three
%dimensional problems, leading to excessively large number of unknowns
%in the linear system. Furthermore, it is rather difficult to obtain
%high-order discretization when the boundary surfaces are irregular.

@@

%The BIE method removes the need of imposing artificial boundary
%conditions for exterior problems and reduces the dimension of the
%problem by one, leading to optimal number of unknowns in the solve
%phase. Very often, one is able to construct the so-called {\it second
%kind integral equation} (SKIE) formulation for these problems, where
%the operator is a sum of the identity operator and a compact operator.
%It is well-known that the condition number of the resulting linear
%system for SKIEs is {\it independent} of the mesh size $h$ or
%equivalently, the number of discretization points. This leads to a
%constant number of iterations when iterative solvers are applied to
%solve the resulting linear system. It further improves the accuracy of
%the solution as the relative error of solution will never increase as
%the number of total discretization points increases. On the other hand,
%the condition number of the linear system from finite difference and
%finite element methods generally increases as a power function of the
%total number of dicretization points, with the exact power depending on
%the underlying spatial dimension and the highest order of the
%differential operator in the PDE. This leads to an increase in the
%number of iterations (and hence the total computational cost) and an
%increase in the relative error of the solution as the number of
%dicretization points increases.

The discretization of the integral operator often leads to a dense
matrix, while finite difference/finite element methods result in sparse
matrices, making matrix-vector product much cheaper to compute.
However,since the invention of the original fast multipole method
(FMM)\cite{fmm5}, there have been many fast
algorithms~\cite{fmm1,fmm2,fmm3,fmm4,fmm6,fmm7,fmm8} that reduce the
cost of matrix-vector product to linear ($O(N)$) or quasilinear
complexity ($O(N \log^kN)$) for dense matrices resulting from the BIE
discretization. Fast direct solvers have also been developed
recently~\cite{fds1,fds2,fds3,fds4,fds5,fds6,fds7,fds8,ho2016cpam2,ho2016cpam1,minden2016,minden2017siammms}.
The development of these fast algorithms have removed a huge hurdle for
the use of boundary integral equations, making the BIE method the method
of choice for constant-coefficient PDEs with linear boundary conditions.
For the exterior Dirichlet problem of the screened Laplace equation, the
SKIE formulation in our previous work~\cite{Fu2018_SIAM} can be readily
extened to the 3D case. For the mobility problem of the Stokes flow in
three dimensions that takes the hydrodynamic effect into account, an
SKIE formulation can be found in~\cite{manasthesis}.

% ----------------------------------------------------------------------
\subsubsection{High-order discretization of surface integrals in three
dimensions}
The practical application of integral equation methods requires the
accurate evaluation of boundary integrals with singular, weakly singular
or nearly singular kernels. There are many numerical quadrature schemes
for dealing with line integrals in two
dimensions~\cite{alpert,kapur,sidi,duffy,bruno1,bruno2,davis_1984,graglia_2008,hackbusch_sauter_1994,
jarvenpaa_2003,khayat_2005,kress_boundary_1991,schwab_1992,
ying_2006,beale1,beale2,goodman_1990, haroldson_1998,
lowengrub_1993,schwab_1992,ggq1,ggq2,ggq3,helsing_2008a,helsing_integral_2009,helsing_tutorial_2012}
and some of them are extremely efficient and can achieve arbitrary high
order.

The high-order quadrature rules for the evaluation of surface integrals
in three dimensions, however, are much less developed than the line
integrals in two dimensions. For example, there are no Gaussian
quadratures for integraing polynomials on a flat triangle, even though
efficient quadratures~\cite{xiao2010cma,vioreanu2014} have been
developed recently for such purpose. For weakly singular or singular
integrals,~\cite{bremer2012jcp,bremer2013jcp} constructed high-order
quadratures for surface integrals on a general triangle,
while~\cite{gimbutas2013sisc} presented a fast algorithm for integrating
$1/r$-type singular integrals for surfaces that are homeomorphic to a
sphere. We plan to study the so-called quadrature by expansion (QBX)
scheme~\cite{klockner2013jcp,qbx2} for the evaluation of both singular
and near-singular surface integrals encountered in the discretization of
BIEs in three dimensions. Conceptually, the idea of the QBX to evaluate
singular, hypersingular and near singular integrals on smooth surfaces
is straightforward. That is, the surface is discretized into smooth
triangles and smooth high-order quadratures are applied to evaluate the
expansion coefficients on all source triangles with the QBX expansion
center placed at a point off the surface. One may then form a suitable
expansion (for example, a Taylor expansion) around that center and
evaluate this expansion back at the target point on the surface (or
close to the surface in the near singular case). Compared to the
competing aforementioned quadrature schemes, the QBX quadrature is
attractive because it offers a clear path for being extended to:
\textbf{(1)} handle any singularity, including hypersingular operators,
\textbf{(2)} be usable with any high-order surface discretization,
\textbf{(3)} generate well-conditioned discrete operators to which
iterative methods such as GMRES~\cite{gmres} can be applied in a
black-box fashion, \textbf{(4)} integrate well with fast algorithms such
as the FMM.

In practice, there are still many issues that need to be resolved. For
example, there are now many variants of QBX including global and local
QBX~\cite{klockner2013jcp,rachh2017jcp}, the target-specific
QBX~\cite{siegel2018jcp}, kernel-independent QBX~\cite{abtin2018bit},
and quadrature by two expansions~\cite{ding2019arxiv}. The coupling of
the QBX and the FMM may also lead to certain instability issues which
may require some changes in the fast multipole
method~\cite{wala2018jcp}. Similar to other quadrature methods, there
have been extensive study on the QBX methods in two dimensions, while
its three dimension
version~\cite{wala2019jcp,af2018sisc,siegel2018jcp,wala2019arxiv} has
not been fully studied and the implementation is even more scarce. We
plan to investigate the accuracy and the convergence order of the
various QBX schemes mentioned above, its coupling with the FMM, parallel
implementation issues for large-scale problems, and the application to
our target problems.

% ----------------------------------------------------------------------
\subsubsection{A hybrid method for close-to-touching Janus particles}
The HAP model for membranes will inevitably lead to the case where some
Janus particles will be very close to each other. Even though QBX can
handle such case accurately, it may lead to excessively large number of
unknowns in order to resolve the unknown densities at those
close-to-touching points. In~\cite{gan2016sisc}, a hybrid scheme is
developed to treat the close-to-touching spheres for the interface
problem of the Laplace equation. The key idea is to combine the analytic
image method and spectrally accurate method of moments, where the
analytic image method will capture the most singular part of the unknown
density at the close-to-touching points. With the aid of the FMM, the
scheme achieves linear complexity with much less number of unknowns on
the material interfaces. We plan to study the hybrid scheme to treat
the close-to-touching case for both the screened Laplace equation and
the mobility problem to further improve the efficiency of the numerical
algorithm for the HAP model.

% ----------------------------------------------------------------------
\subsubsection{Other numerical issues}
Another outstanding numerical issue pertaining to simulating the
self-assembly of amphiphilic particles (such as lipid macromolecles) in
a viscous solvent is the collision between amphiphilic particles. The
hydrophobic attraction potential drives the amphiphilic particles to
move towards each other so to minimize exposure to the solvent. Under
such attraction force the fluid is being squeezed out as particles move
toward each other, and the amphiphilic particles can come into physical
contact in finite time. Such particle collisions in a dense rigid body
suspension is a great challenge and can be a bottleneck in large-scale
simulations. In our previous work a Leonard-Jones potential is used to
keep the particles from overlapping~\cite{Fu2018_SIAM}. Such steep
steric interaction at short ranges introduces great numerical stiffness
that limit the time step size.

Yan {\it et al.} proposed a collision-resolution algorithm within the
framework of boundary integral methods to resolve the non-smooth
many-body dynamics due to collisions by formulating the particle
collision process as a linear complementarity problem with geometric
`non-overlapping' constraint~\cite{Yan2019}. We propose to treat the
particle collisions based on Yan {\it et al.}'s work after the first two
goals (see above) are achieved. A potential complication in combining
the algorithms for particle collision into our QBX integral formulation
is the computational cost for the QBX to refine meshes on particles in
collision.

Once the collision-resolution algorithm is implemented, we will check
the computing performance first before we embark on incorporating
fluctuating hydrodynamics~\cite{Bao17,Bao18}, which might be important
even at the scales of the amphiphilic particle size in some cases. We
propose to extend the high-order time-integration scheme for a
stochastic differential equation~\cite{fu2015pre} to this system.
